---
title: "Introduction to R, part 1c"
author: "Steve Simon"
date: "Monday, August 10, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program briefly introduces the concept of outliers and how you might handle them.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

If you need to, you can clean house and then re-load the data that you created in part1a.Rmd.

```{r load-data}
ls()
rm(list=ls())
load("part1.RData")
ls()
str(fd)
```

There is an unusual data value, which you might not notice right away, but one of the heights is 29.5 inches. We'll talk in more detail about the summary function later, but right now I wanted to show you function because if you have an outlier in your data, you are most likely to discover it by using the summary function.

```{r summary-of-ht}
summary(fd$ht)
```

A height this small is not totally out of the realm of possibility. See, for example, 

--> http://en.wikipedia.org/wiki/List_of_shortest_people

You can use the which function to identify the row with this unusual value for further investigation. Note the use of the double equals sign and how you display a single row of a data frame.

```{r which}
which(fd$ht==29.5)
fd[which(fd$ht==29.5), ]
```

The other values look quite normal. You have to make a careful choice here. One possibility is to do nothing. If you leave the abnormal height in your data set, it may distort some of your graphs and skew some of your statistics. Still, it is often BETTER than some of the alternatives.

A second choice is to remove the entire row from the data frame. The -42 means everything EXCEPT row 42.

```{r remove}
fd1 <- fd[-42,]
fd1$ht
```

A third possibility is to designate the abnormal value as missing. In R, a missing value is represented by NA.

```{r missing}
fd2 <- fd
fd2[42,"ht"] <- NA
fd2$ht
```

There is no one method that is preferred in this setting. If you encounter an unusual value, you should discuss it with your research team, investigate the original data sources, if possible, and review any procedures for handling unusual data values that might be specified in your research protocol.

Your data set may arrive with missing values in it already. Data might be designated as missing for a variety of reasons (lab result lost, value below the limit of detection, patient refused to answer this question) and how you handle missing values is way beyond the scope of this class. Just remember to tread cautiously around missing values as they are a minefield.

Notice that I store the revised data sets with the row removed and with the 29.5 replaced by a missing value in different data frames. This is good programming practice. If you ever have to make a destructive change to your data set (a change that wipes out one or more values or a change that is difficult to undo), it is good form to store the new results in a fresh spot. That way, if you get cold feet, you can easily backtrack.

We'll use the data set with the 29.5 changed to a missing value for all of the remaining analyses of this data set.

## On your own (do this, but you do not need to turn in anything to me)

Display the data for the wt variable and use the summary function. Are there any obvious outliers for weight?

## Save everything

We've created two new data frames, fd1 and fd2, so let's save them before continuing.

```{r save-image}
save.image("part1.RData")
```

Before you go, please answer the following quiz question.

http://www.pmean.com/15/images/quiz1c.html

Send the answers to the quiz by email with the subject line

Introduction to R, (your name), quiz 1c. 
