---
title: "Introduction to R, part 1k"
author: "Steve Simon"
date: "Monday, August 10, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you a sample format for your homework.

I had a student ask, “What, exactly, do I need to turn in for the homework assignments. I said that I’d like to see the R commands, the R output, and a brief commentary. Then I realized that it might be helpful to see a real example. So I made up a simple homework assignment to illustrate what I’m looking for.

1. Read in the data on various body measurements. A description is at
--> http://www.amstat.org/publications/jse/datasets/fat.txt
and the file can be found at 
--> http://www.amstat.org/publications/jse/datasets/fat.dat.txt
Display the first few rows and the last few rows of the data.

2. List all of the bmi values.

3. List the last row of the data set only.

4. Calculate a summary for bmi.

5. Draw a histogram for bmi and interpret it.

6. Show the association between age and bmi using a correlation.

Here's how you would do this using R Markdown and an html format.

Answer to 1.

Here are the R commands to read in the file.

```{r q1}
fn <- "http://www.amstat.org/publications/jse/datasets/fat.dat.txt"
fd <- read.table(file=fn)
```

Here are the first few rows of the data set.

```{r q1a}
head(fd)
```
Here are the last few rows of the data set.

```{r q1b}
tail(fd)
```

Answer to 2. 

Here are all the bmi values.

```{r q2}
names(fd) <- c("case","fat.b","fat.s","dens","age","wt","ht","bmi","ffw","neck","chest","abdomen","hip","thigh","knee","ankle","biceps","forearm","wrist")
fd$bmi
```

Answer to 3.

Here is the last row of the data.

```{r q3}
fd[252, ]
```

Answer to 4. 

Here is a summary for bmi.

```{r q4}
summary(fd$bmi)
```

[[NOTE: INTERPRETATION IS IMPORTANT, BOTH HERE AND FOR THE NEXT TWO QUESTIONS. You don't have to describe every single number, just pick two or three to talk about.]] Rounding to two significant figures, the mean is 25, the range is 18 to 49.

Answer to 5. 

Here is a histogram for bmi.

```{r q5}
hist(fd$bmi)
```

The distribution is slightly skewed with a possible outlier between 45 and 50.

Answer to 6.

Here is a correlation between bmi and age.

```{r q6}
cor(fd$bmi, fd$age)
```

The correlation is 0.12. There is little or no association between bmi and age.

Concluding remarks. What you see here is only a suggestion. Feel free to create a format that you like. The three things I am looking for is a listing of the R commands that you used, the text or graphical output, and a brief interpretation.