---
title: "Introduction to R, part 1h"
author: "Steve Simon"
date: "Monday, August 10, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to compute correlations.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

If you need to, you can clean house and then re-load the data that you created earlier.

```{r load-data}
ls()
rm(list=ls())
load("part1.RData")
ls()
```

The correlation coefficient is a single number between -1 and +1 that quantifies the strength and direction of a relationship between two continuous variables. As a rough rule of thumb, a correlation larger than +0.7 indicates a strong positive association and a correlation smaller than -0.7 indicates a strong negative association. A correlation between +0.3 and +0.7 (-0.3 and -0.7) indicates a weak positive (negative) association. A correlation between -0.3 and +0.3 indicates little or no association.

Don't take these rules too literally. You're not trying to make definitive statements about your data set. You are just trying to get comfortable with some general patterns that occur in your data set. A complex and definitive statistical analysis will almost certainly not agree with at least some of the preliminary correlations noted here.

```{r corr-1}
cor(fd2$fat.b,fd2$age)
```

## Correlation matrices

You can get a matrix of correlations for every possible pair of variables. This command becomes a bit more complicated if there are some categorical variables in your data set, as you need to exclude these prior to calculating the correlation matrix. Since you are just trying to get a general feel for your data, a bit of rounding will help you. Also, you should remove the case number before calculating the correlation matrix.

```{r corr-2}
cor(fd2[ ,-1])
round(cor(fd2[,-1]),2)
round(100*cor(fd2[,-1]))
```

You can get pieces of the correlation matrix by supplying pieces of the data frame.

```{r corr-3}
leg.measures <- c("hip","thigh","knee","ankle")
arm.measures <- c("biceps","forearm","wrist")
round(100*cor(fd2[,leg.measures],fd2[,arm.measures]))
```

We created a few new objects, so you might want to save things here before proceeding to the next section.

```{r save-everything}
save.image("part1.RData")
```

Before you go, please answer the following quiz question.

1. What function in R is used to calculate a correlation?

Please also take this opportunity to ask any questions you might have.

Send the answers to the quiz by email with the subject line

Introduction to R, (your name), quiz 1h. 
