---
title: "Introduction to R, part 1d"
author: "Steve Simon"
date: "Monday, August 10, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to calculate some simple summary statistics.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

If you need to, you can clean house and then re-load the data that you created earlier.

```{r load-data}
ls()
rm(list=ls())
load("part1.RData")
ls()
```

When you first encounter any data set, you should run some simple univariate statistics and graphics on any column of data that you plan to use in any of your analyses. In this data set, all the variables are continuous, and the summary function works well.

```{r summary}
summary(fd2$fat.b)
summary(fd2$ht)
```

## On your own (do this, but you do not need to turn in anything to me)

Calculate a summary for bmi.

What happens if you ask for a summary of the entire data frame? (summary(fd2)).

## Save everything we've done so far

```{r save-image}
save.image("part1.RData")
```

Before you go, please answer the following quiz question.

1. What statistics are produced by the summary function if your data is continuous?

Please also take this opportunity to ask any questions you might have.

Send the answers to the quiz by email with the subject line

Introduction to R, (your name), quiz 1d. 
