---
title: "Introduction to R, part 1e"
author: "Steve Simon"
date: "Monday, August 10, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you some basic things you need to know about missing value codes in R.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

If you need to, you can clean house and then re-load the data that you created earlier.

```{r load-data}
ls()
rm(list=ls())
load("part1.RData")
ls()
```

Notice that the summary function for the ht variable notes that one of the values is missing. You should watch these missing values obsessively. This can get a bit tricky. 

```{r tracking-missing-1}
which(fd2$ht==NA)
```

Logic involving missing values is tricky. If you checking for equality and one of the things in the comparison is missing, then the result is neither TRUE, not FALSE, but rather missing.

Fair enough, but R takes it a bit further, and if both sides when you are checking for equality are missing, then they might both be 5 is they weren't missing or maybe one might be 5 and the other one 10. So it might be TRUE or it might be FALSE, so we're better off calling the logical result as missing.

This is called a three valued logic system and it has advantages and disadvantages. I won't get into any technical details, except to say that you should never make assumptions. Check what you do when you are working with missing values to make sure that the three valued logic system doesn't produce results that you didn't expect.

The short term solution to the above problem is to use a special function, is.na.

```{r tracking-missing-2}
which(is.na(fd2$ht))
```

The summary function will trap and remove missing values, but most other functions in R will, by default, report a result as missing if any values going into that function are missing. The philosophy in R, I suppose, is that you need to actively select an approach for handling missing values rather than relying on a lazy default. Also, a  missing value is less likely to evade your notice when you have to actively specify an approach for handling the missing values.

```{r mean-sd-1}
mean(fd2$ht)
sd(fd2$ht)
```

Read the help file for these functions (enter ?mean or ?sd at the command prompt).

Look carefully and note that the na.rm option allows you to compute the statistic after missing values are removed.

```{r mean-sd-2}
mean(fd2$ht,na.rm=TRUE)
sd(fd2$ht,na.rm=TRUE)
```

For univariate functions, there are only two realistic ways to handle missing values, but for bivariate and multivariate function, there are a multitude of approaches, such as pairwise deletion, listwise deletion, last observation carried forward, single imputation, and multiple imputation. There is a lot of controversy over various methods for handling missing values.

You haven't created any new data frames, but save eveything anyway.

```{r save-image}
save.image("part1.RData")
```

Before you go, please answer the following quiz question.

http://www.pmean.com/15/images/quiz1e.html

Please also take this opportunity to ask any questions you might have.

Send the answers to the quiz by email with the subject line

Introduction to R, (your name), quiz 1e. 

