---
title: "Introduction to R, part 1b"
author: "Steve Simon"
date: "Monday, August 10, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to display parts of your data frame.

```{r version-and-current-date}
R.version.string
Sys.Date()
```

To take up where you left off, you may need to load the data that you saved earlier. If you have not interrupted your work, you may not need to do this, but it costs nothing other than a few electrons and a few seconds to run this. The ls function shows what objects you currently have available and the str function shows information about a particular object. You probably haven't forgetten what you stored in your data frame, but sometimes, you drop a projects for days or even weeks and when you return, you need a bit of a refresher.

```{r load-data}
ls()
rm(list=ls())
load("part1.RData")
ls()
str(fd)
```

To list the entire data frame, just type the name. In R, when you type the name of an object, it knows that you want a display of what is in that object. To list just part of a data frame, you can use a matrix style of notation.

```{r single-column}
fd
fd[1, 1] # first row, first column
fd[1, ] # all columns for first row
fd[, 1] # all rows for first column
fd[1, "ht"] # first row, "ht" column
fd[1:50, "ht"] # first fifty rows, "ht" column
```

There is a short cut for displaying a particular column that uses the $ sign. Notice how the subset of fd$ht differs from the subset of the entire data frame.

```{r dollar-sign}
fd$ht # all rows for "ht" column
fd$ht[1:50] # first fifty rows for "ht" column
# Notice that fd$ht[1:50] and fd[1:50, "ht"] produce the same results.
```

## On your own (do this, but you do not need to turn in anything to me)

Display the first ten rows and the first five columns of your data.

## Save everything

Are you ready for another break? You did not change anything, so there's no need to save anything. But it doesn't hurt to get in the habit of saving regularly.

```{r save-image}
save.image("part1.RData")
```

Before you go, please answer the following quiz question.

http://www.pmean.com/15/images/quiz1b.html

Send the answers to the quiz by email with the subject line

Introduction to R, (your name), quiz 1b.
