---
title: "Introduction to R, part 1i"
author: "Steve Simon"
date: "Monday, August 10, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to draw scatterplots.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

If you need to, you can clean house and then re-load the data that you created earlier.

```{r load-data}
ls()
rm(list=ls())
load("part1.RData")
ls()
```

he best graphical way to display a relationship between two continuous variables is a scatterplot.

```{r scatterplot-1}
plot(fd2$abdomen,fd2$fat.b)
plot(fd2$ht,fd2$abdomen)
```

There are lots of options available to customize your graph. Here are just a few. The xlab and ylab arguments in the plot function control what is displayed on the horizontal (x) and vertical (y) axes. The pch argument control what is used as the plotting character.

```{r scatterplot-2}
plot(fd2$abdomen/2.54,fd2$fat.b,
     xlab="Abdomen circumference in inches",
     ylab="Percentage of body fat using Brozek's equation",
     pch="+")
```

Notice that you can do the units conversion "on the fly". This doesn't
work on all functions.

## Adding trend lines

These scatterpltos often benefit by adding a trend line.

Here's a linear regression trend line

```{r trend-1}
fd2$abdomen.in <- fd2$abdomen/2.54
trend.line <- lm(fat.b~abdomen.in,data=fd2)
plot(fd2$abdomen.in,fd2$fat.b)
abline(trend.line)
```

Here's a smooth curve.

```{r trend-2}
smooth.curve <- lowess(fd2$age,fd2$fat.b)
plot(fd2$age,fd2$fat.b)
lines(smooth.curve)
```

## Try all of these things on a totally different data set.

There is a second data set on sleep in mammals. You can find a brief description of this data set at

--> http://www.statsci.org/data/general/sleep.html

and you can download the actual data at

--> http://www.statsci.org/data/general/sleep.txt

Notice that the last three variables are ordinal, but for the purposes of this class, we will treat them the same as the other variables.

## Read the file into R

Notice that this file has variable names at the very top. Look at the help file for read.table and see what option you might choose here. Display the first few rows and the last few rows.

As noted in each of the R commands shown below, I will remove the answers before I distribute this file to the students.

```{r read-sleep}
fn <- "http://www.statsci.org/data/general/sleep.txt"
sl <- read.table(file=fn,header=TRUE)
head(sl)
tail(sl)
```

## Your homework assignment for Part 1.

1. Notice that there is a huge range in body weight. Display the information for the smallest and the largest mammals.

2. Which variables have missing data?

3. Calculate the mean and standard deviation for TotalSleep.

4. Draw a histogram for the BodyWt variable. Note that this variable is highly skewed. Re-draw the histogram on the log scale. Hint: log(sl$BodyWt,base=10). For exta credit, relabel the axes with values at 0.001,0.01, etc.

5. Calculate the ratio of BrainWt to BodyWt and express it as a percentage. Be sure to convert grams to kilograms (or vice versa) before computing the ratio. What animal has the smallest/largest ratio?

6. Do bigger animals sleep longer or live longer? Show your results using a correlation coefficient.

7. Do animals who have high degrees of Predation, Exposure, or Danger sleep less? Show your results using a plot and a trendline.

8. Place all the key results into a Word document, add a bit of commentary, and turn it in by email. Use the subject line: "Introduction to R, Part 1 Homework"

## If there is time.

If there is time, a third data set is described at

--> http://www.amstat.org/publications/jse/datasets/body.txt

and the data itself is at

--> http://www.amstat.org/publications/jse/datasets/body.dat.txt

We created a few new objects, so you might want to save things here before proceeding to the next section.

```{r save-everything}
save.image("part1.RData")
```

There is no quiz for part 1i.