---
title: "Introduction to R, part 1j"
author: "Steve Simon"
date: "Monday, August 10, 2016"
output: html_document
---

This program is part of a series of R programs for the Introduction to R class. This program shows you how to import a wider range of file type.

```{r r-version-and-current-date}
R.version.string
Sys.Date()
```

If you need to, you can clean house, but there is no need to re-load the data that you created earlier.

```{r load-data}
ls()
rm(list=ls())
```

The material for Part 1 did not include enough information about getting data into and out of R. Here are a few extra examples.

## Read in comma separated values

The read.csv function will read a text file with commas separating individual data points. This is a commonly used format and frequently has the extension .csv.

In notepad or another text editor, type in the following
values.

1,4

2,8

3,12

4,16

and save the file using the name simple.csv. Note the directory that you are saving it in.

Here's a time saving hint: run the getwd function without any arguments (getwd()) to see where the current working directory. This is the place that R is currently looking for files if you don't provide a path. If you store the file simple.csv in the current working directory, then you don't have to type in the path name as well as the filename.

You can use the setwd command to change the current working directory. Be sure to use forward slashes (/) in your pathname rather than backslashes (\).

```{r read-csv}
fn <- "~/introduction-to-r-part1/dat/simple.csv"
d1 <- read.csv(file=fn)
d1
```

Oops! That didn't work the way I wanted it to. Look at the help file for read.csv. What option would you use to avoid losing the first line of data?

## On your own  (do this, but you do not need to turn in anything to me)

Add the line x,y to the top of your csv file and read it in using the header=TRUE option.

Note: depending on how you typed your data, you may or may not get an error message like "Warning in readLines(file, n = thisblock): incomplete final line found"

This happens when R does not find a carriage return at the end of the last line of data. This is easily fixed by opening the text file and going to the very end with CTRL-END.

if you land at the end of the very last line, you don't have a final carriage return. If you land on a blank line below the last line of text, then you are fine.

## Using the more general function, read.table

Review the options for the read.table command.

Type in the same data, but using the tilde (~) as a separator (e.g., 1~4). Save the file as tilde.txt.

Here is how you would use read.table to read this file.

```{r read-table}
fn <- "tilde.txt"
d2 <- read.table(file=fn,header=TRUE,sep="~")
```

## Reading in space delimited files.

Open up notepad and type in the following lines.

1  4

2  8

3 12

4 16

Be sure to line up the numbers properly by inserting two blanks between 1 and 4 and two blanks between 2 and 8, but only one blank between 3 and 12 and between 4 and 16.

Save the file as space.txt.

If you tried to read in the file with sep=" ", you would get an error. Try this to see what the error looks like.


You can either always edit the file and replace any double blanks with single blanks, but if you look carefully at the help file, you will see that the argument sep="" (which is the default) will use white space (one or more blanks, tabs, carriage returns, or newlines). This works pretty well but not always. We will see a file where you have to fix things before you can import it properly.

```{r read-space}
fn <- "space.txt"
# Try read.table(fn,sep=" ") first.
d3 <- read.table(fn,sep="")
```

Note that the default for read.table is header=FALSE, as opposed to the default of HEADER=TRUE for read.csv.

## Reading fixed width format files

Sometimes you will get a file with no delimiters to save space. This requires that each variable takes up a fixed number of columns and that information is often specified in a separate file.

Create a fixed width file in notepad with the following lines.

1 4

2 8

312

416

and store it as fixed.txt.

```{r fixed}
fn <- "fixed.txt"
d4 <- read.fwf(fn,c(1,2))
d4
```

You should also consider the read.fortran function. Check out the help file.

## Writing text files

If you want to write a text file, you can use either the write.csv or write.table functions. Look up the help files on these functions. We'll create a small data frame and store it in a comma delimited and tab delimited format.

```{r write-text}
d5 <- data.frame(x=c(13,12,11),y=c(2,4,6))
fn <- "~/introduction-to-r-part1/results/output1.csv"
write.csv(d5,fn)
fn <- "~/introduction-to-r-part1/results/output2.txt"
write.table(d5,fn)
```

Open these files in notepad to see what they look like.

## On your own (do this, but you do not need to turn it anything to me)

To avoid the extra line at the top and the row names "1","2","3" use the options header=FALSE and col.names=FALSE. Try this and see what the two files look like now.

## Writing binary files

By default, R saves everything that you create on exiting from the program and stores it in a file with no name and with an extension of RData (in other words, .RData). This file is automatically loaded the next time you start R from the same subdirectory.

If you want to save a data frame in binary format, use the function save(). Binary formats are much faster but are pretty much useful only within R itself.

In other words, don't send a binary file to someone who does not have R. Binary files are useful when you are moving data from one R user to another, or data from one computer to another, or sometimes when you are moving data from one folder to another.

We will not be using any of these files again, so there is no need to save anything.

There is no quiz for part 1j.


